Detection and replication of epistasis influencing transcription in humans

### 2013-06-18 docs/manuscript/nature_letter.tex
gib and joseph final changes to ms

\begin{abstract}
Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following stringent Bonferroni correction for multiple testing. Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects ($p < 1.0 \times 10^{-16}$). There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans.

@@

% \subsection{Introduction}
In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently, and cumulatively to the mean of a trait. This framework has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}. Outside the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.

@@

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood. After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing 238 gene expression levels. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).

Additionally, there was enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication (Figure \ref{fig:qqMeta}). We observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Supplementary Figure \ref{fig:qqMetaNonsig}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication. It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. Therefore these results are very encouraging with regards to the detection and replication of epistasis.

% \subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection. The majority of our discovery interactions comprised of one SNP that had a previously known association and one SNP that had no previous association in the dataset \cite{Powell2013} (439 out of 501). Only 9 interactions were between SNPs that both had marginal effects while 64 were between SNPs that had no known marginal effects. Additionally, we observed that the largest epistatic variance component for the 501 interactions was divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by chance ($p = 0.22$ for divergence from expectation). This is not surprising because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.

@@

% \subsection{Functional annotation}
In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment (Methods). We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}). There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing). However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues). This difference between \emph{cis} and \emph{trans} SNPs suggests a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions. There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}). This enrichment is consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$). In particular, there was substantial enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs, a known regulator of transcription in haematopoietic cells \cite{Tsai1994} ($p = 1 \times 10^{-40}$).

We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.

@@

However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components. Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes. By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance. It must be noted that there are at least three caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different effect sizes. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.

@@

\subsection{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes. Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}. Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone. All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$). Of these, 434 passed filtering in both replication datasets. A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{Grey_graph_of_interactions_2_lists.pdf}
	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes). 316 Interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 118 interactions that did not replicate are depicted in grey. It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
	\label{fig:fireworks}
\end{figure}
\clearpage

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{gpbonfrep.pdf}
	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class. Phenotypes are for gene transcript levels (dark coloured tiles = low expression, light coloured tiles = high expression). Columns of GP maps are for each independent population. Rows of GP maps are for each of 30 significantly replicated interactions, corresponding to the rows in Table \ref{tab:bonferroni}. There is a general trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}
\clearpage

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{qqMeta.pdf}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} Top panel shows all 434 discovery SNPs that were tested for interactions. Observed $p$-values ($y$-axis, $-\log_{10}$ scale) are plotted against the expected $p$-values ($x$-axis, $-\log_{10}$ scale). The multiple testing correction threshold for significance is denoted by a dotted line. Bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets. The shaded grey area represents the 5\% confidence interval for the expected distribution of $p$-values. Dark blue points represent $p$-values that exceed the confidence interval, light blue are within the confidence interval.}
	\label{fig:qqMeta}
\end{figure}
\clearpage

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{qqMetaNonsig}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets, excluding the 30 points significant at the Bonferroni level} The right panel (Null) shows the interaction $p$-values from a meta analysis across two independent datasets on 434 randomly drawn SNP pairs. The left panel (Empirical) shows the interaction $p$-values from the 404 putative interactions that were not significant at the Bonferroni correction threshold. Dark blue points represent $p$-values that surpass the 5\% FDR level, as in Figure \ref{fig:qqMeta}.}
\label{fig:qqMetaNonsig}
\end{figure}
\clearpage

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{501_pairs_snp_list_Grid}
	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3 \cite{Trynka2013}. Enrichment $p$-values are calculated using permutation analysis for 34 different cell types ($y$-axis) in four tissues (Rows of boxes). There is enrichment for \emph{cis}-acting SNPs in Haematopoeitic tissue types only. \emph{Trans}-acting SNPs have no tissue specificity.}
	\label{fig:h3k4me3}
\end{figure}
\clearpage

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{chromosome_interactions}
	\caption{\textbf{Number of overlaps between chromosome interactions and epistatic interactions} Interacting chromosome regions may be a possible mechanism underlying epistatic interactions. The number of epistatic interactions within 20kb, 500kb, 2Mb and 10Mb of known chromosome interacting regions are shown by red vertical lines. The histograms represent the null distribution based on random sampling of 10000 datasets for each window size.}
	\label{fig:chromosomeinteractions}
\end{figure}
\clearpage

\begin{figure}
	\centering
	\includegraphics[width=5in]{enhancers}
	\caption{\textbf{There is enrichment for enhancer sequences for \emph{cis} and \emph{trans} SNPs}}
	\label{fig:enhancers}
\end{figure}

>>>>

\begin{abstract}
Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Three hundred and forty-five interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} at a false dicovery rate (FDR) of 2.5\%, with 30 significant at a conservative $p < 0.05$ Bonferroni level. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans.

@@

% \subsection{Introduction}
In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently and cumulatively to the mean of a trait. This framework has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}. Outside the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.

@@

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood. After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).

In addition, we observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication. It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. Therefore these results are encouraging with regards to the detection and replication of epistasis.

% \subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection. The majority of our discovery interactions comprised of one SNP that had a previously known association and one SNP that had no previous association in the dataset \cite{Powell2013} (439 out of 501). Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects. Additionally, we observed that the largest epistatic variance component for the 501 interactions was divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by chance ($p = 0.22$ for divergence from expectation). This is not surprising because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.

@@

% \subsection{Functional annotation}
In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment (Methods). We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}). There was significant enrichment for \emph{cis}-acting SNPs in haematopoietic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing). However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues). This difference between \emph{cis} and \emph{trans} SNPs suggests a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions. There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a}, consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$). In particular, there was enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs, a known regulator of transcription in haematopoietic cells \cite{Tsai1994} ($p = 1 \times 10^{-40}$).

We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.

@@

However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components. Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes. By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance (Methods). It must be noted that there are at least three caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different effect sizes. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.

@@

\subsection{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7,339 probes representing 6,158 RefSeq genes. Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}. Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone. All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$). Of these, 434 passed filtering in both replication datasets. A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{Grey_graph_of_interactions_2_lists.pdf}
	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes). Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 89 interactions that did not replicate are depicted in grey. It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
	\label{fig:fireworks}
\end{figure}
\clearpage

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{gpbonfrep.pdf}
	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class. Phenotypes are for gene transcript levels (dark coloured tiles = low expression, light coloured tiles = high expression). Columns of GP maps are for each independent population. Rows of GP maps are for each of 30 significantly replicated interactions at the Bonferroni level, corresponding to the rows in Table \ref{tab:bonferroni}. There is a clear trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}
\clearpage

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{qqMeta.pdf}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions. Observed $p$-values ($y$-axis, $-\log_{10}$ scale) are plotted against the expected $p$-values ($x$-axis, $-\log_{10}$ scale). The multiple testing correction threshold for significance following Bonferroni correction is denoted by a dotted line. The bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets. The shaded grey area represents the 5\% confidence interval for the expected distribution of $p$-values. Dark blue points represent $p$-values that exceed the confidence interval, light blue are within the confidence interval.}
	\label{fig:qqMeta}
\end{figure}
\clearpage

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{qqMetaNonsig}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets, excluding the 30 points significant at the Bonferroni level} The right panel (Null) shows the interaction $p$-values from a meta analysis across two independent datasets on 434 randomly drawn SNP pairs. The left panel (Empirical) shows the interaction $p$-values from the 404 putative interactions that were not significant at the Bonferroni correction threshold. Dark blue points represent $p$-values that surpass the 2.5\% FDR level, as in Figure \ref{fig:qqMeta}.}
\label{fig:qqMetaNonsig}
\end{figure}
\clearpage

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{501_pairs_snp_list_Grid}
	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3 \cite{Trynka2013}. Enrichment $p$-values are calculated using permutation analysis for 34 different cell types ($y$-axis) in four tissue types (Rows of boxes). There is enrichment for \emph{cis}-acting SNPs in Haematopoietic tissue types only. \emph{Trans}-acting SNPs have no tissue specificity.}
	\label{fig:h3k4me3}
\end{figure}
\clearpage

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{chromosome_interactions}
	\caption{\textbf{Number of overlaps between chromosome interactions and epistatic interactions} Interacting chromosome regions may be a possible mechanism underlying epistatic interactions. The number of epistatic interactions within 20kb, 500kb, 2Mb and 10Mb of known chromosome interacting regions are shown by red vertical lines. The histograms represent the null distribution based on random sampling of 10,000 datasets for each window size.}
	\label{fig:chromosomeinteractions}
\end{figure}
\clearpage
